Beyond Static Retrieval: Adaptive Method for Improving RAG Performance

### 2025-03-17 Paper/latex/paper_RAG.tex
update

Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.Despite their strengths,they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering.However,(RAG) faces several limitations. For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally,A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}. often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes K a key factor in balancing performance and efficiency.

Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches like those proposed yuan et al.\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval. However, these approaches face significant limitations, including a lack of adaptability, high computational overhead, and a dependency on high-quality metadata for optimal performance. These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.

@@

This paper contributes to the existing literature by reviewing the various approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static, dynamic, and Hybrid selection strategies. Additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance. On account of this, there arises a pressing need for an improved approach capable of addressing the challenges of both static and dynamic in retrieval-augmented systems. \\
The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.\\
The following section,introduces the novel hybrid dynamic selection algorithm. Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing static and dynamic approaches. Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities

@@

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries. For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.

Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval :combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} retrieval framework designed to improve the relevance between queries and documents. In the first stage, a similarity matching method retrieves an initial set of documents. In the second stage, a deeper analysis is performed to identify dynamic-relevant documents, supported by a classifier that filters out irrelevant ones . however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.

Our method leverages adaptive thresholds to dynamically refine the list of retrieved documents, to make sure that only the most contextually relevant matches are included. In contrast to traditional static or fixed K-selection techniques, which employ a consistent retrieval strategy for every query, our approach adapts to the specific characteristics of both the query and the dataset. This flexibility allows the algorithm to better accommodate diverse contexts, enhancing the relevance and quality of retrieval results.

@@

Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}, dense retrieval\cite{karpukhin2020dpr}, and others. For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
1. Encoding each object into a specific representation.\\
2. Building an index to organize the data for efficient searching .

@@

Once the retriever fetches relevant data from external sources,the generator processes and integrates this information into a well-structured and meaningful answer. At the core of this process is a Large Language Model (LLM), which ensures that the generated text is not only fluent and accurate but also remains relevant to the original query.

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

@@

\subsection{Retrieval Algorithm}
We propose an \textbf{adaptive threshold mechanism} (referred to as (Dynamic Candidate Selection) that leverages the Mixture of Logits (MoL)framework to enhance the candidate retrieval process. The mechanism operates as follows:

@@

The query q represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework, q is transformed into Pq
embeddings through a multi-layer perceptron (MLP). While our hybrid algorithm does not explicitly employ an MLP for this transformation
\subsection{Hyperparameter settings.}
For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU. We use the Adam optimizer with a learning rate of 0.001. For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in table \ref{tab:100k_movies} ,table \ref{tab:1m_movies}

@@

\begin{table}[h]
	\centering
		\large
	\renewcommand{\arraystretch}{1.3}
	\begin{tabular}{|l|c|c|c|c|c|c|}
		\hline
		\textbf{Model} & \textbf{HR@1} & \textbf{HR@10} & \textbf{HR@50} & \textbf{HR@200} & \textbf{MRR} & \textbf{NDCG} \\ \hline
		\multicolumn{7}{|c|}{\textbf{MovieLens 100K}} \\ \hline
		SASRec+Hybrid & 0.0070 & 0.0775 & 0.2606 & 0.5704 & 0.0363 & 0.1241 \\ \hline
		SASRec+MoL    & 0.0000 & 0.0775 & 0.2887 & 0.5775 & 0.0285 & 0.1182 \\ \hline
		\multicolumn{7}{|c|}{\textbf{MovieLens 1M}} \\ \hline
		SASRec+Hybrid & 0.0464 & 0.1810 & 0.3819 & 0.6159 & 0.0925 & 0.1829 \\ \hline
		SASRec+MoL    & 0.0508 & 0.1799 & 0.3896 & 0.6126 & 0.0971 & 0.1859 \\ \hline
	\end{tabular}
			\caption{Performance Comparison of SASRec+Hybrid and SASRec+MoL on MovieLens 100K and 1M Datasets}
				\label{tab:results_tab}
\end{table}
\newpage
On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}. The hybrid algorithm particularly excels in early position recommendations, as evidenced by the non-zero HR@1 score.

@@

Choosing the right value for K in retrieval-based systems is a significant challenge, as it directly impacts both the efficiency and accuracy of information retrieval. While using a fixed K is straightforward, it often fails to adapt to the varying complexities of different queries. This can result in either missing important information or retrieving too much irrelevant data, which adds noise to the results. On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.
To address these limitations, we introduced a hybrid dynamic K selection strategy that strikes a balance between adaptability and efficiency. Our approach adjusts K based on the unique characteristics of each query while keeping computational demands manageable. This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.
Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.

>>>>

Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.Despite their strengths,they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations. For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally,A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}. often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes (k) a key factor in balancing performance and efficiency.

Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval. However, these approaches face significant limitations, including , high computational overhead, and a dependency on high-quality metadata for optimal performance. These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.

@@

This paper contributes to the existing literature by reviewing the various approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static, dynamic, and Hybrid selection strategies. Additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance. On account of this, there arises a pressing need for an improved approach capable of addressing the challenges of both static and dynamic in retrieval-augmented systems. \\
The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.\\
The following section,introduces the novel hybrid dynamic selection algorithm. Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing  approaches. Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.

@@

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks..

Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal (k) for retrieval, A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust (k) by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} is a two-stage retrieval framework.first, it retrieves an initial set of (k1) documents followed by a classifier that evaluates the dynamic relevance between documents to enhance recall and retrieve additional relevant 
(k2) documents, however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.

Our method leverages adaptive thresholds to dynamically refine the list of retrieved documents, to make sure that only the most contextually relevant matches are included. In contrast to traditional static or fixed K-selection techniques, which employ a consistent retrieval strategy for every query, our approach adapts to the specific characteristics of both the query and the dataset.

@@

Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}which relies on exact term matching techniques like BM25 and TF-IDF, dense retrieval\cite{karpukhin2020dpr}which Uses deep learning models to encode queries and documents into dense vector embeddings, retrieving relevant documents based on semantic similarity rather than exact word matching, and others. For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
1. Encoding each object into a specific representation (e.g., term-based for sparse retrieval, vector embeddings for dense retrieval).\\
2. Building an index to organize and efficiently retrieve relevant data during search .

@@

Once the retriever fetches relevant data from external sources,the generator processes and integrates this information into a well-structured and meaningful  response. At the core of this process is a Large Language Model (LLM), which ensures that the generated text is not only fluent and accurate but also remains relevant to the original query.

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

@@

\subsection{Retrieval Algorithm}
We propose an \textbf{adaptive threshold mechanism} (referred to as (Dynamic Candidate Selection) that leverages the Mixture of Logits (MoL) framework to enhance the candidate retrieval process. The mechanism operates as follows:

@@

The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
embeddings through a multi-layer perceptron (MLP).
\subsection{Hyperparameter settings.}
For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU. We use the Adam optimizer with a learning rate of 0.001. For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in table \ref{tab:100k_movies} ,table \ref{tab:1m_movies}

@@

\begin{table}[h]
	\centering
		\large
	\renewcommand{\arraystretch}{1.3}
	\begin{tabular}{|l|c|c|c|c|c|c|}
		\hline
		\textbf{Model} & \textbf{HR@1} & \textbf{HR@10} & \textbf{HR@50} & \textbf{HR@200} & \textbf{MRR} & \textbf{NDCG} \\ \hline
		\multicolumn{7}{|c|}{\textbf{MovieLens 100K}} \\ \hline
		SASRec+Hybrid & 0.0070 & 0.0775 & 0.2606 & 0.5704 & 0.0363 & 0.1241 \\ \hline
		SASRec+MoL    & 0.0000 & 0.0775 & 0.2887 & 0.5775 & 0.0285 & 0.1182 \\ \hline
		\multicolumn{7}{|c|}{\textbf{MovieLens 1M}} \\ \hline
		SASRec+Hybrid & 0.0464 & 0.1810 & 0.3819 & 0.6159 & 0.0925 & 0.1829 \\ \hline
		SASRec+MoL    & 0.0508 & 0.1799 & 0.3896 & 0.6126 & 0.0971 & 0.1859 \\ \hline
	\end{tabular}
			\caption{Performance Comparison of SASRec+Hybrid and SASRec+MoL on MovieLens 100K and 1M Datasets}
				\label{tab:results_tab}
\end{table}

On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}. The hybrid algorithm particularly excels in early position recommendations, as evidenced by the non-zero HR@1 score.

@@

Choosing the right value for K in retrieval-based systems is a significant challenge, as it directly impacts both the efficiency and accuracy of information retrieval. While using a fixed K is straightforward, it often fails to adapt to the varying complexities of different queries. This can result in either missing important information or retrieving irrelevant data, which adds noise to the results. On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.
To address these limitations, we introduced a hybrid dynamic K selection strategy that strikes a balance between adaptability and efficiency. Our approach adjusts K based on the unique characteristics of each query while keeping computational demands manageable. This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.
Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
